---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# radsafer

The goal of radsafer is to provide functions that are useful for radiation safety professionals.

## Installation

You can install the released version of radsafer from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("radsafer")
```
Or install the development version from GitHub:
 
```{r eval = FALSE}
# install.packages("devtools")
devtools::install_github("markhogue/radsafer")
```

## Oveview
To start using the installed package: 

```{r}
library(radsafer)
```

### Decay Correction Functions
Radsafer includes several functions to manage radioactive decay corrections:

`dk_cf` provides a correction factor. Revise a calibration or source check value to today's date (the default) or a date and time of your choosing. 

```{r}
dk_cf(half_life = 5.27, date1 = "2010-12-01", date2 = "2018-12-01", time_unit = "y")
```
Use this function to correct for the value needed today. Say, a disk source originally had a target count rate of 3000 cpm:
```{r}
3000 * dk_cf(half_life = 5.27, date1 = "2010-12-01", date2 = "2018-12-01", time_unit = "y")
```

Other decay functions answer the following questions: 
* What is the decayed activity? `dk_activity`, Given a percentage reduction in activity, how many half-lives have passed.`dk_pct_to_num_half_life`

* How long will it take to reach the target radioactivity? `dk_time`

* Given the radioactivity at one time, what was the radioactivity at an earlier time? `dk_reverse`

* Given two data points, estimate the half-life: `half_life_2pt`

### Instrumentation functions
Correct _vented ion chamber readings_ based on difference in air pressure (readings in degrees Celsius and mm Hg):
```{r}
air_dens_cf(T.actual = 30, P.actual = 760, T.ref = 20, P.ref = 760)
```

Correct for _geometry_ when reading a close _neutron_ source. Example: neutron rem detector with a radius of 11 cm and source near surface:
```{r}
neutron_geom_cf(11.1, 11)
```

_Scaler counts_: obtain quick distributions for parameters of interest:
```{r}
 scaler_sim(true_bkg = 50, true_samp = 10, ct_time = 20, trials = 1e5)
```

_Rate meters_: Ratemeter readings are plotted once per second for 600 seconds. The meter starts with a reading of zero and builds up based on the time constant. Resolution uncertainty is established to express the uncertainty from reading an analog scale, including the instability of its readings. Many standard references identify the precision or resolution uncertainty of analog readings as half of the smallest increment. This should be considered the single coverage uncertainty for a very stable reading. When a reading is not very stable, evaluation of the reading fluctuation is evaluated in terms of numbers of scale increments covered by meter indication over a reasonable evaluation period. Example with default time constant:
```{r}
rate_meter_sim(cpm_equilibrium = 270, meter_scale_increments = seq(100, 1000, 20))
```
To estimate _time constant_, use `tau.estimate`

### Stay-time computation
Given a dose rate, dose allowed, and a safety margin (default = 20%), calculate stay time with: `stay_time`
```{r}
stay_time(dose_rate = 120, dose_allowed = 100, margin =  20)
```

### MCNP utility functions
If you create MCNP inputs, these functions may be helpful:

* Create an _energy distribution_ from histogram data with: `si_hist` and `sp_hist` (Load the data into R first using copy and paste with `scan` or reading from an external table with, for example, `read.table`.)

* Determine the entries needed for MCNP _coordinate transformation rotation_ with `rot_fun`

* Quickly obtain the _cone angle_ entry with `tan_d2`

* For _MCNP outputs_, plot the results of a tally with _energy bins_:
```{r echo = FALSE}
plot_data <- photons_cs137_hist[1:16, ]
```
```{r}
plot_spec_from_hist(plot_data, "mR_h", "Example: Cs-137 with scatter")
```
