---
title: "Introduction to radsafer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intro_and_decay_family}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  )
```

```{r setup}
library(radsafer)
```

The radsafer package was developed with these goals:

* Provide functions that are commonly used in radiation safety

* Provide easy access to data provided in the RadData package

* Share some less commonly-used functions that may be of significant value to workers in instrumentation and modeling

Related functions are identified as members of thes families:

* decay corrections 

* radionuclides 

* rad measurements 

* mcnp tools 

This framework makes it a little easier to find the function you want - the help file for each function lists related functions.

### Decay Correction Functions
Radsafer includes several functions to manage radioactive decay corrections:

**dk_cf** provides a correction factor. Revise a calibration or source check value to today's date (the default) or a date and time of your choosing. 

```{r}
dk_cf(half_life = 5.27, date1 = "2010-12-01", date2 = "2018-12-01", time_unit = "y")
```
*By the way, answers in R are typically provided on screen with an indexing system, in case there are lots of numbers to keep track of. So, the [1] above indicates that this is the first answer in our string of, well, only one answer.* 

In this example, there is a Co-60 source that was calibrated on December 1, 2010, but we need to know how much to correct for radioactive decay (half of the activity decaying away every 5.27 years) to the date of interest, December 1, 2018. Using it this way gives you a numerical result that is the needed correction factor. Next, we see a more direct way to use this result.

Here's an example where a calibration source originally provided a target count rate of 3000 cpm on our instrument. To find out the corrected count rate, we multiply the original count rate by this correction factor:
```{r}
3000 * dk_cf(half_life = 5.27, date1 = "2010-12-01", date2 = "2018-12-01", time_unit = "y")
```

Other decay functions answer the following questions: 
* What is the decayed activity? **dk_activity**, Given a percentage reduction in activity, how many half-lives have passed.**dk_pct_to_num_half_life**

* How long will it take to reach the target radioactivity? **dk_time**

* Given the radioactivity at one time, what was the radioactivity at an earlier time? **dk_reverse** 

* Given two data points, estimate the half-life: **half_life_2pt**

### rad measurements functions
**air_dens_cf**
Correct _vented ion chamber readings_ based on difference in air pressure (readings in degrees Celsius and mm Hg):
```{r}
air_dens_cf(T.actual = 30, P.actual = 760, T.ref = 20, P.ref = 760)
```
Let's try it out combined with the instrument reading:
```{r}
rdg <- 100
(rdg_corrected <- rdg * air_dens_cf(T.actual = 30, P.actual = 760, T.ref = 20, P.ref = 760))
```

**neutron_geom_cf**

Correct for _geometry_ when reading a close _neutron_ source. Example: neutron rem detector with a radius of 11 cm and source near surface:
```{r}
neutron_geom_cf(11.1, 11)
```

**disk_to_disk_solid_angle**

Correct for a mismatch between the _source calibration_ of a _counting system_ and the item being measured. A significant factor in the counting efficiency is the solid angle from the source to the detector. You can also check for the impact of an item not being centered with the detector.

Example: You are counting an air sample with an active collection diameter of 45 mm, your detector has a radius of 25 mm and there is a gap between the two of 5 mm. (The function is based on radius, not diameter so be sure to divide the diameter by two.) The relative solid angle is:

```{r}
(as_rel_solid_angle <- as.numeric(disk_to_disk_solid_angle(r.source = 45/2, gap = 20, r.detector = 12.5, runs = 1e4, plot.opt = "n")))
```

An optional plot is available in 2D or 3D:
```{r out.width = '50%', fig.align='center'}
(as_rel_solid_angle <- as.numeric(disk_to_disk_solid_angle(r.source = 45/2, gap = 20, r.detector = 12.5, runs = 1e4, plot.opt = "3d")))
```
 
Continuing the example: the only calibration source you had available with the appropriate isotope has an active diameter of 20 mm. Is this a big deal? Let's estimate the relative solid angle of the calibration, then take a ratio of the two.

```{r}
(cal_rel_solid_angle <- disk_to_disk_solid_angle(r.source = 20, gap = 20, r.detector = 12.5, runs = 1e4, plot.opt = "n"))
```
Correct for the mismatch:
```{r}
(cf <- cal_rel_solid_angle / as_rel_solid_angle)
```
This makes sense - the air sample has particles originating outside the source radius, so more of them will be lost, thus an adjustment is needed for the activity measurement.

**scaler_sim**

_Scaler counts_: obtain quick distributions for parameters of interest:
```{r out.width = '50%', fig.align='center'}
scaler_sim(true_bkg = 50, true_samp = 10, ct_time = 20, trials = 1e5)
```

**rate_meter_sim**

_Rate meters_: In the ratemeter simulation, readings are plotted once per second for a default time of 600 seconds. The meter starts with a reading of zero and builds up based on the time constant. Resolution uncertainty is established to express the uncertainty from reading an analog scale, including the instability of its readings. Many standard references identify the precision or resolution uncertainty of analog readings as half of the smallest increment. This should be considered the single coverage uncertainty for a very stable reading. When a reading is not very stable, evaluation of the reading fluctuation is evaluated in terms of numbers of scale increments covered by meter indication over a reasonable evaluation period. Example with default time constant:
```{r out.width = '50%', fig.align='center'}
rate_meter_sim(cpm_equilibrium = 270, meter_scale_increments = seq(100, 1000, 20))
```

To estimate _time constant_, use `tau.estimate`

### Stay-time computation

Given a dose rate, dose allowed, and a safety margin (default = 20%), calculate stay time with: `stay_time`
```{r}
stay_time(dose_rate = 120, dose_allowed = 100, margin =  20)
```

### mcnp tools functions

If you create MCNP inputs, these functions may be helpful:

**mcnp_si_sp_RD**
Obtain emission data from the RadData package and write to a file for use with the radiation transport code, MCNP.

**mcnp_si_sp_hist** 

* Create an _energy distribution_ from histogram data with: `mcnp_si_sp_hist` Options are provided to use data already in R and to copy and paste data interactively.

**mcnp_matrix_rotations**

* Determine the entries needed for MCNP _coordinate transformation rotation_ 

**mcnp_cone_angle**

* Quickly obtain the _cone angle_ entry

**mcnp_plot_out_spec**

For _MCNP outputs_, plot the results of a tally with _energy bins_. Either first save your data to a text file, or copy and paste it using `mcnp_scan2spec.df`. Then plot it using your favorite method, or do a quick plot with `mcnp_plot_out_spec`:

```{r echo = FALSE}
plot_data <- photons_cs137_hist[1:16, ]
```
```{r out.width = '50%', fig.align='center', warning=FALSE, message=FALSE}
mcnp_plot_out_spec(photons_cs137_hist, 'example Cs-137 well irradiator')
```

### radionuclides
Search by alpha, beta, photon or use the general screen option. 

`search_phot_by_E` allows screening based on energy, half-life, and minimum probability. Also available are `search_alpha_by_E`, `search_beta_by_E`, and `bin_screen_phot`. `bin_screen_phot` allows limiting searches to radionuclides with emissions in an energy bin of interest with additional filters for not having photons in other specified energy bins. Results for all these search functions may be plotted with `RN_plt`.

Here's a search for photon energy between 0.99 and 1.01 MeV, half-life between 13 and 15 minutes, and probability at least 1e-4

```{r echo = TRUE}
search_results <- search_phot_by_E(0.99, 1.01, 13 * 60, 15 * 60, 1e-4)
```

```{r echo = FALSE}
knitr::kable(search_results)
```

```{r echo = TRUE, fig.align='center', fig.width=4}
RN_plt(spec_0.1_0.3)
```

The `RN_index_screen` function helps find a radionuclide of interest based on decay mode, half-life, and total emission energy.

In this example, we search for radionuclides decaying by spontaneous fission with half-lives between 6 months and 2 years.

```{r echo = TRUE}
RNs_selected <- RN_index_screen(dk_mode = "SF", min_half_life_seconds = 0.5 * 3.153e7, max_half_life_seconds = 2 * 3.153e7)
```

```{r echo = FALSE}
knitr::kable(RNs_selected[, c(1:3)])
```

Other radionuclides family functions provide specific activity and short tables of decay data.
